Designing Types for R, Empirically

### 2020-05-14 paper/main.tex
before refocus

Another possibility is that a value is used {\it without being forced}, as
is the case when values are ``metaprogrammed''.
This is common practice in e.g. the \code{tidyverse} ecosystem, as they use the metaprogramming capabilities of R to implement a DSL for data analysis.

To tackle this, we make an {\it initial guess} of function argument types on
function entry. This collects types for argument default values, and
collects an ``unused'' type for unevaluated expressions.  If an argument
promise is later forced, we simply update the recorded type for the
argument.

\item Second, we wanted to account for formal arguments (arguments which are
  named in the function definition) that were {\it not} passed to the
  function. This represents cases where a function was not called with all of its
  arguments, and no default value was specified for missing arguments. 
  We record a ``missing'' type for such argument.

There are two obvious ways to deal with missing arguments: type them as bot
or as top.  As we are performing a dynamic analysis based on package test
code, we conservatively type these arguments as top, or \code{any}, as It is
impossible to know dynamically what all possible function behaviours are.
We discuss the prevalence of such any-typed arguments in Section~\ref{sec:evaluation}.

\item Third, we had to contend with R's C implementation sometimes
  performing longjumps, which discards current R calls from the call stack.
  In order to ensure that call traces are not lost when a longjump occurs,
  \typetracer intercepts the unwinding process initiated by the longjump and
  mimics the functionality of the functions being exited.  The R-dyntrace
  framework supplies the return value during the unwinding process (if there
  is indeed such a value), so we can capture it, analyze it, and include it
  in the call trace.  In certain situations, however, no return value can be
  obtained (e.g. a longjump occurred as part of some error handling), and
  then we simply record a ``jumped'' return value, which we can deal with in
  post-processing.  
  % These longjumps occur quite often, particularly when S3 dispatch occurs, as the R implementation optimizes returns from dispatched functions by initiating a longjump.

@@

As mentioned, our analysis is built on the R-dyntrace framework.  R-dyntrace
is a very efficient instrumentation framework for R, which is built on a
modified version of the R runtime.  Our dynamic type analysis only needs to
be run once, ever---from the analysis results, we synthesize function
signatures which are passed along to later type-checking tools which do not
rely on a heavily modified runtime.

We primarily rely on 8 R-dyntrace callbacks: \texttt{closure\_entry},
\texttt{closure\_exit}, \texttt{builtin\_entry}, \texttt{builtin\_exit},
\texttt{special\_entry},\texttt{special\_exit},
\texttt{promise\_force\_entry}, and \texttt{promise\_force\_exit}.  These
callbacks fire when closures, builtin, and special functions are entered and
exited, and before and after the forcing of promises.  The function-related
callbacks are used mainly for bookkeeping: the analysis is notified that a
closure, builtin, or special has been entered/exited by pushing/popping the
call onto a stack.  The calls themselves store a {\it call trace} object,
where we store the type information that we collect, populated with initial
guesses of argument types (as discussed in
Section~\ref{subsec:typetracer-challenges}).  Further, the type of dispatch
is recorded on the {\tt \_entry} variants: recall that R can perform single or multiple dispatch on function arguments depending on their class.
We collect dispatch information in order to analyze dispatch usage patterns.

The main work is done in {\tt promise\_force\_exit}.  As arguments to
functions in R are wrapped in promises, we delay our reflection to promise
force time.  In R, promises are essentially objects with value and
expression slots.  When a promise is created for a particular expression,
that expression is merely put into the expression slot of a promise object,
and a special {\it unbound} value is loaded into the value slot (unless some
default value is specified, wherein it is loaded into the value slot---this
is the case with default argument values).  When the promise is forced, the
expression is evaluated in the forcing context, and the value is stored in
the value slot of the promise.  So to get the type of a function argument,
we check the value slot of the argument promise---perhaps recursively if the
value itself is a promise.

@@

The type information we collect is far more detailed than the types we
actually generate, and so the next stage of our inference process is to
distill the rich type information into simpler, manageable types.  For each
unique observed call trace, we translate the representation of the type of
each of the function arguments and returns according to the design of our
type system: for instance, we might have determined that a function returned
a vector of integer with 4 elements, with names, that was free of NA values,
in which case we would write out \code{integer[]} as the type.

@@

@@

\contractr is an R package that implements our type system for the R
package ecosystem. It adds contracts to package functions for type-checking
arguments and results on function calls. On contract failure, it reports a type
mismatch warning with the package name, function name, argument name, argument
position, expected type signature, actual type signature of the value observed,
and a stack trace for the called function. \contractr works by modifying
function definition to insert a call to the type-checking function. When the
modified function is called, the type-checking function is invoked on the
modified function's arguments and return value.

\contractr is implemented in 434 R LOC and 2358 C++ LOC. The core
implementation is in C++ to keep the runtime overhead low. It been designed and
tested with GNU R-3.5.0 but it also works with recent versions of R. It has been
hardened with a battery of 400 test cases. We have used it extensively during
the course of this work; firstly, for sanity checking of XXX type signatures
generated for the XXX packages during the development phase, and secondly, for
assessing the quality of these type signatures on XXX packages during the
evaluation phase.

@@

This example also illustrates an issue we grappled with when dealing with structs.
On line 8, we see the call \code{data(sweetpotato)}.
It is quite common for R example code to use some built-in data sets, such as \code{sweetpotato}, to show off some package functionality with ``real data''.
We found that the types of functions which use these built-in data sets (such as \code{aov}) became polluted with the names from the example sets.

@@

When \typetracer is configured to detect structs we saw a marked increase in misrepresentative types.
We implemented a unification strategy for struct types, removing them if they co-occurred with list types, and turning them into lists if two co-occurring struct types shared no common names, but even so we found ourselves with function signatures that were full of noise introduced by the struct names.

We still believe struct types to be useful to R programmers, and so our language supports the type and \contractr will check them, though we configured \typetracer to ignore them as they made the data substantially less useful.
On the positive side, we found that \AT{get the number, but often} structs-typed values had a {\it class}, and so we decided to include class-based types.
Classes are discussed in more detail in the next section, though we will quickly discuss data frames before focusing on them.

@@

One way to deal with data frames is through the struct type, with a named field for each column of the data frame, but as mentioned previously structs introduced undue noise into \typetracer's analysis results. 
Further complicated issues for data frames is that many functions built to operate on them operate in a name-agnostic way.
For instance, the \code{tidyverse} package ecosystem allows programmers to pass column names to functions which operate on their data frames.
In base R, idiomatic data frame use is to use string column names to select rows from the frame (unless only a single column is of interest, wherein the \code{$} syntax is appropriate). 
We include some examples in Figure~\ref{fig:data-frames-bad} to illustrate the complexities.

@@

\begin{lstlisting}
# Let's put some examples showing that data frames can get out of hand, and that
# structs are inadequate.
\end{lstlisting}

@@

Doing justice to this type will be an endeavour in and of itself, requiring reconciliation of base R's \code{data.frame}, \code{tidyverse}'s \code{tibble}, and \code{data.table}'s \code{data.table}.
\AT{We could talk about how metaprogramming is pervasive in dplyr, and how we would need to capture that, too, which is a huge undertaking.}
It will also require proper treatment of function types, and perhaps even parametric types, which dynamic analyses such as \typetracer are unable to discern soundly.
\AT{General words to convince the reader that data frames are, in fact, hard, and worthy of their own paper.}

@@

%
%
%
%
\subsection{Classes in R}

@@

%
%
\subsubsection{Select Examples}

>>>>

Another possibility is that a value is used {\it without being forced}, as
is the case when values are ``metaprogrammed''.
This is common practice in e.g. the \code{tidyverse}~\footnote{\AT{either talk about it here or in background}} ecosystem, as they use the metaprogramming capabilities of R to implement a DSL for data analysis.

To tackle possibly unevaluated arguments, we make an {\it initial guess} of function argument types on
function entry. Specifically, we collect types for argument default values, and
ascribe an ``unused'' type for unevaluated expressions.  If an argument
promise is later forced, we simply update the recorded type for the
argument.

\item Second, we wanted to account for formal arguments (arguments which are
  named in the function definition) which received no values when the function was called. This represents cases where a function was not called with all of its
  arguments, and no default value was specified for missing arguments. 
  We record a ``missing'' type for such argument.

There are two obvious ways to deal with missing arguments: type them as $\bot$
or as $\top$.  As we are performing a dynamic analysis based on package test
code, we conservatively type these arguments as $\top$, or \code{any}, as It is
impossible to know dynamically what all possible function behaviours are.
We discuss the prevalence of such any-typed arguments in Section~\ref{sec:evaluation}.

\item Third, we had to contend with R's backend sometimes
  performing longjumps, which discards current R calls from the call stack.
  In order to ensure that call traces are not lost when a longjump occurs,
  \typetracer intercepts the unwinding process initiated by the longjump and
  mimics the functionality of the functions being exited.  The R-dyntrace
  framework supplies the return value during the unwinding process (if there
  is indeed such a value), so we can capture it, analyze it, and include it
  in the call trace.  In certain situations, however, no return value can be
  obtained (e.g. a longjump occurred as part of some error handling), and
  then we simply record a ``jumped'' return value, which we can deal with in
  post-processing.  
  % These longjumps occur quite often, particularly when S3 dispatch occurs, as the R implementation optimizes returns from dispatched functions by initiating a longjump.

@@

As mentioned, our analysis is built on the R-dyntrace framework.  R-dyntrace
is a very efficient instrumentation framework for R, which is built on a
modified version of the R runtime. 
% Our dynamic type analysis only needs to be run once, ever---from the analysis results, we synthesize function signatures which are passed along to later type-checking tools which do not rely on a heavily modified runtime.

We primarily rely on 8 R-dyntrace callbacks: \texttt{closure\_entry},
\texttt{closure\_exit}, \texttt{builtin\_entry}, \texttt{builtin\_exit},
\texttt{special\_entry},\texttt{special\_exit},
\texttt{promise\_force\_entry}, and \texttt{promise\_force\_exit}.  These
callbacks fire when closures, builtin, and special functions are entered and
exited, and before and after the forcing of promises.  The function-related
callbacks are used mainly for bookkeeping: the analysis is notified that a
closure, builtin, or special has been entered/exited by pushing/popping the
call onto a stack.  The calls themselves store a {\it call trace} object,
where we store the type information that we collect, populated with initial
guesses of argument types (as discussed in
Section~\ref{subsec:typetracer-challenges}).  
Also, recall that R can perform single or multiple dispatch on function arguments depending on their class.
We collect such dispatch information the {\tt \_entry} variants.

The primary functionality of \typetracer is in {\tt promise\_force\_exit}.  As arguments to
functions in R are wrapped in promises, we delay our reflection to promise
force time.  In R, promises are essentially objects with value and
expression slots. \AT{Put this as part of R-dyntrace background:} When a promise is created for a particular expression,
that expression is merely put into the expression slot of a promise object,
and a special {\it unbound} value is loaded into the value slot (unless some
default value is specified, wherein it is loaded into the value slot---this
is the case with default argument values).  When the promise is forced, the
expression is evaluated in the forcing context, and the value is stored in
the value slot of the promise.  So to get the type of a function argument,
we check the value slot of the argument promise---recursively if the
value itself is a promise.

@@

\AT{Maybe further discuss how we can create many candidate TS from the data.}

The type information we collect is far more detailed than the types we
actually generate (which we discuss in Section~\ref{sec:types}), and so the next stage of our inference process is to
distill the rich type information into simpler, manageable types.  For each
unique observed call trace, we translate the representation of the type of
each of the function arguments and returns according to the design of our
type system: for instance, we might have determined that a function returned
a vector of integer with 4 elements, with names, that was free of NA values,
in which case we would write out \code{integer[]} as the type.

@@

Thus far, we have discussed \typetracer and our strategies to take the data from the dynamic analysis and generate function types.
This work also presents a framework for specifying these types in R code, as well as generating contract checks for these types.
We dub this tool \contractr, discussed next.

@@

\contractr is an R package that implements our type system for the R
package ecosystem. For functions with type annotations, it adds contracts for type-checking
arguments and results on function calls. On contract failure, it reports a type
mismatch warning with the package name, function name, argument name, argument
position, expected type signature, actual type signature of the value observed,
and a stack trace for the called function. \contractr works by modifying
function definition to insert a call to the type-checking function. When the
modified function is called, the type-checking function is invoked on the
modified function's arguments and return value.

\contractr is implemented in 434 R LOC and 2358 C++ LOC. The core
implementation is in C++ to keep the runtime overhead low. It been designed and
tested with GNU R-3.5.0 but it also works with recent versions of R. It has been
hardened with a battery of 400 test cases. We have used it extensively during
the course of this work; firstly, for sanity checking of XXX type signatures
generated for the XXX packages during the development phase, and secondly, for
assessing the quality of these type signatures on XXX packages during the
evaluation phase. We discuss this further in Section~\ref{sec:evaluation}.

@@

Unfortunately, the inclusion of structs complicated our analysis by polluting the data, illustrated on line 8 with the call \code{data(sweetpotato)}.
It is quite common for R example code to use some built-in data sets, such as \code{sweetpotato}, to show off some package functionality with ``real data''.
When \typetracer is configured to detect structs we saw a marked increase in misrepresentative types, as the types of functions which use these data sets (such as \code{aov} in Figure~\ref{fig:struct-ex}) became polluted with the names from the example sets.
We implemented a unification strategy for struct types, removing them if they co-occurred with list types, and turning them into lists if two co-occurring struct types shared no common names, but even so we found ourselves with function signatures that were full of noise introduced by the struct names.

We still believe struct types to be useful to R programmers, and so our language supports the type and \contractr will check them, though we configured \typetracer to ignore them as they made the data substantially less useful.
On the positive side, we found that \AT{get the number, but often} structs-typed values had a {\it class}, which helped motivate our inclusion of class-like types in our type annotation language.
Classes are discussed in more detail in Section~\ref{subsec:classes}.

@@

@@

One way to deal with data frames is through the struct type, with a named field for each column of the data frame, but as mentioned previously structs introduced undue noise into \typetracer's analysis results. 
Further complicating data frames is that many functions built to operate on them operate in a name-agnostic way.
For instance, the \code{tidyverse} package ecosystem allows programmers to pass column names to functions which operate on their data frames.
In base R, idiomatic data frame use is to use string column names to select rows from the frame (unless only a single column is of interest, wherein the \code{$} syntax is appropriate). 
We include some examples in Figure~\ref{fig:data-frames-bad} to illustrate the complexities.

@@

\begin{lstlisting}
data(cars)

# tidyverse: Filter rows where speed is > 20. 
fast_cars_tidy <- filter(cars, speed > 20)

# base R: Get all rows where the "speed" column has a value > 20.
fast_cars_base <- cars[cars[, "speed"] > 20, ] # base example

\end{lstlisting}

@@

Doing justice to this type will be an endeavour in and of itself, requiring reconciliation of base R's \code{data.frame}, \code{tidyverse}'s \code{tibble}, and \code{data.table}'s \code{data.table}.
Developing useful types for the \code{tidyverse} ecosystem may well involve

It will also require proper treatment of function types, and perhaps even parametric types, which dynamic analyses such as \typetracer are unable to discern soundly.
\AT{General words to convince the reader that data frames are, in fact, hard, and worthy of their own paper.}

@@

%
%
%
%
\subsection{Classes in R}
\label{subsec:classes}

@@

%
%
\subsubsection{Subsumed Checks}
